# Logistic regression

```{r}
date()
```

```{r}
library(tidyverse)
alc <- read_csv("./data/alc.csv")
dim(alc)
colnames(alc)
```

## Short description of dataset 
Dataset is created from data: https://archive.ics.uci.edu/ml/datasets/Student+Performance  
The data approaches student achievement in secondary education of two Portuguese schools, and the dataset joins two student alcohol consumption datasets. The variables not used for joining the two data have been combined by averaging (including the grade variables). 'alc_use' is the average of 'Dalc' and 'Walc'. 'high_use' is TRUE if 'alc_use' is higher than 2 and FALSE otherwise.    
  
Dataset alc contains 35 variables with 370 observations.  


## My personal hypothesis about relationships of  with alcohol consumption
I think interesting variables in relation to alcohol consumption are sex, number of past class failures, final grade (G3) and number of school absences.  
My hypothesis for these variables are:  
1. Men's alcohol consumption is higher than women's.  
2. Those who consume more alcohol have more past class failures compared to those who consume less alcohol.   
3. Those who consume more alcohol have lower final grades compared to those who consume less alcohol.  
4. Those who consume more alcohol have more school absences compared to those who consume less alcohol.    
    
```{r}
library(dplyr); library(ggplot2); library(GGally); library(finalfit)


dependent <- "high_use"
explanatory <- c("sex", "failures", "G3", "absences")
alc %>% 
  summary_factorlist(dependent, explanatory, p = TRUE,
                     add_dependent_label = TRUE)

#median and quartiles of the numeric variables
summary(alc$failures)
summary(alc$G3)
summary(alc$absences)

#draw a barplot of alcohol consumption by sex
ggplot(data = alc, aes(x = high_use, fill=sex)) + geom_bar()

#produce summary statistics of final grade by alcohol consumption and sex
alc %>% group_by(sex, high_use) %>% summarise(mean_grade=mean(G3),count = n())

#draw a boxplot of final grade by alcohol consumption and sex
ggplot(alc, aes(x = high_use, y = G3, col=sex)) + geom_boxplot() + ylab("grade")

#produce summary statistics of school abscences by alcohol consumption and sex
alc %>% group_by(sex, high_use) %>% summarise(mean_absences=mean(absences),count = n())

#draw a boxplot of school absences by alcohol consumption and sex
ggplot(alc, aes(x = high_use, y = absences, col=sex)) + geom_boxplot() + ylab("absences") + ggtitle("Student absences by alcohol consumption and sex")
```  
  
## Description of chosen variables and their relationships with alcohol consumption  
There are 195 women and 175 men. Of those, 21% (41/195) and 40% (70/175) have high alcohol consumption, respectively. This supports my hypothesis of men consuming more alcohol than women.   
  
The mean of past class failures is 0.2 (range 0-3). Of high users, 3% (3/111) have 3 failures, 8% (9/111) have 2 failures, and 11% have 1 failure. Of others, 0.4% (1/259) have 3 failures, 3% (8/259) have 2 failures, and 5% have 1 failure. Thus it would seem that students with high alcohol consumption have failed class more often than those with low alcohol consumption.  
  
The mean final grade is 11.5 (range 0-18). Men with high alcohol consumption seem to have lower final grades than people with low alcohol consumption. Interestingly, women with high alcohol consumption seem to have around the same final grades than people with low alcohol consumption. So my hypothesis is supported only by observations of men.   
  
The mean number of absences is 4.5 (range 0-45, median 3). The median of school absences is higher for men with high alcohol consumption than other groups. However, if we look at the mean, women with high alcohol consumption have the highest mean of absences, and both men and women with high alcohol consumption have higher mean than those with low alcohol consumption. This is somewhat in line with my hypothesis.   
   
    
```{r}
# fit a logistic regression model 
my_model <- glm(high_use ~ sex + failures + absences + G3, data = alc, family = "binomial")

# print out a summary of the model
summary(my_model)

# fit a new model without G3 model 
my_model2 <- glm(high_use ~ sex + failures + absences, data = alc, family = "binomial")

# print out a summary of the model
summary(my_model2)

# compute odds ratios (OR)
OR <- coef(my_model2) %>% exp

# compute confidence intervals (CI)
CI <- confint(my_model2) %>% exp

# print out the odds ratios with their confidence intervals
cbind(OR, CI)

```  
   
## Logistic regression
In the first model, final grade has a p-value 0.2, and thus isn't associated with high alcohol consumption. Let's fit the model again without it. In the second model, all variables are associated with the outcome, and also the AIC drops from 415.59 to 414.99, which indicates that G3 can be dropped from the model.  
  
According to this model, high alcohol consumption is associated with being male (OR 2.7, 95% CI 1.7-4.4, p-value <0.001), having failed class (OR 1.8, 95% CI 1.2-2.8, p-value 0.004), and school absences (OR 1.1, 95% CI 1.1-1.2, p-value <0.001).  
My primary hypothesis were correct regarding being male and having failed classes. However, there is only slight association with school absence (OR 1.1), and the final grade doesn't seem to be associated with high alcohol consumption.   
    
```{r}
# predict() the probability of high_use
probabilities <- predict(my_model2, type = "response")

library(dplyr)
# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prediction = probability>0.5)

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction)

# draw a plot of 'high_use' versus 'probability' in 'alc'
ggplot(alc, aes(x = probability, y = high_use)) + geom_point()

# define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, prob = alc$probability)
```  
   
## Predictive power of the model
In the 2x2 table we can see that 252 low and 33 high alcohol consumers are predicted correctly. In the predictions, however, there are 78 false negatives and 7 false positives. According to the training error, 23% of the individuals are classified inaccurately. This is better than tossing a coin (where the chances would be 50/50), but I guess with some background information one could get quite close with educated guesses.  
 
```{r} 
# 10-fold cross-validation
library(boot)
cv <- cv.glm(data = alc, cost = loss_func, glmfit = my_model2, K = 10)

# average number of wrong predictions in the cross validation
cv$delta[1]

#Let's try the same with another model. 
my_model3 <- glm(high_use ~ traveltime + sex + absences +studytime + goout, data = alc, family = "binomial")
summary(my_model3)
prob2 <- predict(my_model3, type = "response")
loss_func(class = alc$high_use, prob = prob2)
cv2 <- cv.glm(data = alc, cost = loss_func, glmfit = my_model3, K = 10)
cv2$delta[1]
``` 
    
## 10-fold cross-validation of the model  
The prediction error of the model when using 10-fold cross-validation is around 0.24, which is slightly less than in the exercise set.  
With the other tested model (explanatory variables being sex, home to school travel time, number of school absences, weekly study time, and going out with friends), the prediction error with 10-fold cross-validation is around 0.22, which is still lower.  
  
  
  

