# Regression and model validation

```{r}
date()
```
  
## Basic information about dataset   
  
```{r}
library(tidyverse)
students2014 <- read_csv("./data/learning2014.csv")
dim(students2014)
str(students2014)
```
  
Dataset is created from data: https://www.mv.helsinki.fi/home/kvehkala/JYTmooc/JYTOPKYS3-meta.txt  
Data is collected from students in course 'Johdatus yhteiskuntatilastotieteeseen' in fall 2014. 

Dataset students2014 contains 7 variables with 166 observations: gender, age, attitude, deep, stra, surf, and points.

All variables in the original data related to strategic, deep, and surface learning have been combined into three variables (stra, deep, and surf, respectively), and scaled to original scales by taking the mean.

Attitude tells about student's global attitude towards statistics. Variable has been scaled back to the original scale of the questions, by dividing it with the number of questions.

Points tells the course exam points. Observations with points equaling 0 have been excluded, since those students didn't attend the course exam. 
  
## Describing the data  
  
```{r}
library(GGally)
library(ggplot2)

# create a plot matrix with ggpairs()
p <- ggpairs(students2014, mapping = aes(col=gender, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))
p

summary(factor(students2014$gender))
summary(students2014$age)
summary(students2014$attitude)
summary(students2014$deep)
summary(students2014$stra)
summary(students2014$surf)
summary(students2014$points)
```
  
There are 66% (110/166) women in respondents.
The median age is 22 years (range 17-55), the median age of women seems to be a bit lower than that of men.

The median attitude is 3.2 (range 1.4-5). The median attitude of men seems higher than that of women.

The median for deep learning is 3.7 (range 1.6-4.9).
The median for strategic learning is 3.1 (range 1.3-5).
The median for surface learning is 2.8 (range 1.6-4.3).
For women, the median for strategic and surface learning seems a bit higher than for men. 

The median for exam points is 23 (range 7-33). The median is about the same in men and women. 

Attitude correlates with exam points, so that with "better" attitude towards statistics one usually gets higher exam points.

Surface learning correlates negatively with deep and strategic learning as well as attitude. So with better attitude one less often utilizes surface learning. If one utilizes surface learning, one less often utilizes deep or strategic learning. 
  
## Interpreting the fitted regression model  
  
```{r}
#fit a linear regression model using points as the outcome and attitude, strategig and surface learning as explanatory variables
my_model <- lm(points ~ attitude + stra + surf, data = students2014)
summary(my_model)

#remove surface learning from the model, since it's so far from significant
my_model2 <- lm(points ~ attitude + stra, data = students2014)
summary(my_model2)
#error rate
sigma(my_model2)*100/mean(students2014$points)
```
  
First we choose three variables that correlate the most with points in above visualization (namely attitude, strategic and surface learning). However, after fitting the model, we see that for surface learning there is a really high p-value (0.466), so we remove that and fit the model again with just attitude and strategic learning. 

In the new model, p-value of the F-statistic is 7.734e-09. So at least one of the explanatory variables is significantly associated with exam points.

For attitude, p-value is <0.001, so it is significantly associated with exam points. According to this model, if attitude increases by 1, exam points increase by approximately 3.5. 

For strategic learning, p-value is 0.089. It isn't <0.05 that's usually deemed significant, however, it is quite close. So we probably don't want to state that strategic learning has no effect on exam points. On the other hand, the estimated intercept for strategic learning is 0.9, so a change in attitude makes a much bigger difference than a change in strategic learning. 

The adjusted R-squared is 0.1951, which means that attitude (and strategic learning) only explain around 20% of the variance in exam points. 

The residual standard error is 5.289 which corresponds to 23% prediction error rate. 
  
## Assumptions of the model  
  
```{r}
plot(my_model2, which=c(1,2,5))
```
  
Linear regression assumes a linear relationship between predictors and outcome, that residual errors are normally distributed, that residuals have a constant variance, and independence of residual error terms. 

With **Residuals vs. Fitted** plot we can check the linearity of data. Here, the red line seems to be horizontal at approximately zero, where it should be. So we can assume a linear relationship. 

With plot **Normal Q-Q** we can check if residual errors are normally distributed. The plot of residuals should more or less follow the dashed line. Here they mostly do so, with some exceptions at the upper and lower ends. However, we may assume normality.

The **Residuals vs. Leverage** plot marks three most extreme points (35, 71, 145). Two of them exceed 3 standard deviations, so they are possible outliers.