# Analysis of longitudinal data  
  
```{r}
date()
```  
  
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(lme4)
```  
  
## Analysis of Longitudinal Data I: Graphical Displays and Summary Measure Approach  
  
```{r}
RATSL <- read.table("./data/RATSL.txt", header=TRUE)
# Factor variables ID and Group
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

# Glimpse the data
glimpse(RATSL)

# Check the column names
names(RATSL)

# Look at the structure of dataset
str(RATSL)

# Print out summaries of the variables
summary(RATSL)
```  
  
The dataset RATSL is a long format version of data from a nutrition study conducted in three groups of rats. The groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately) weekly, except in week seven when two recordings were taken) over a 9-week period.  
  
The dataset has 176 rows and 5 columns (ID, group, weight, and day in borth character and integer formats).  
  
## Plot the RATSL data    
```{r}
# Plot the data
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype=Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") + 
  theme(legend.position = "top")

```  
  
## Summary graph  
  
```{r}
# Calculate the weightgain
RATSL <- 
  RATSL %>%
  group_by(ID) %>%
  mutate(Weightgain = Weight - Weight[Time == 1] ) %>%
  ungroup()

# Plot the weightgain
ggplot(RATSL, aes(x = Time, y = Weightgain, group = ID)) +
  geom_line(aes(linetype=Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weightgain (grams)") + 
  theme(legend.position = "top")

# Summary data with mean and standard error of weightgain by group and time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weightgain), se = sd(Weightgain) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(Weightgain) +/- se(Weightgain)")

```  
  
Based on the graph, group 2 seems to have gained most weight in the study period.    
  
## Checking for outliers  
  
```{r}
# Draw a boxplot of the mean weightgain versus group
ggplot(RATSL, aes(x = Group, y = Weightgain)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean weightgain, days 1-64")
```  
  
No outliers detected.  
  
##  T-test and Anova  
  
```{r}
# Pairwise t-test comparisons
pwc <- RATSL %>%
  pairwise_t_test(Weightgain ~ Group, p.adjust.method = "bonferroni")
pwc

# Visualization: box plots with p-values
pwc <- pwc %>% add_xy_position(x = "Group", step.increase = 1)
ggboxplot(RATSL, x = "Group", y = "Weightgain") +
  stat_pvalue_manual(pwc, hide.ns = TRUE) 

# Fit the linear model with the mean as the response 
fit <- lm(Weightgain~Group, data = RATSL)
summary(fit)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```    
  
Based on the pairwise t-tests and anova, there seems to be a significant difference in weight gain between groups 1 and 2, and 2 and 3.  
  
## Analysis of Longitudinal Data II: Linear Mixed Effects Models for Normal Response Variables  
  
```{r}
BPRSL <- read.table("./data/BPRSL.txt", header=TRUE)
# Factor treatment & subject
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

# Glimpse the data
glimpse(BPRSL)

# Check the column names
names(BPRSL)

# Look at the structure of dataset
str(BPRSL)

# Print out summaries of the variables
summary(BPRSL)
```  
  
The dataset BPRSL is a long format version of BPRS, in which 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.  
  
The dataset has 360 rows and 5 columns (treatment, subject, bprs, and week in both character and integer formats).   
  
I ran out of time for the interpretations of this second part.    
  
## Plot the BPRLS data  
  
```{r}  
# Plot the data
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))  
```  
  
## The linear model  
  
```{r}
# create a regression model BPRS_reg
BPRS_reg <- lm(bprs~week+treatment, data=BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```  
  
## The random intercept model  
  
```{r}
# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```  
  
## Random intercept and random slope model  
  
```{r}
# create a random intercept and random slope model
BPPS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPPS_ref1)

# perform an ANOVA test on the two models
anova(BPPS_ref1, BPRS_ref)
```  
  
## Random intercept and random slope model with interaction  
  
```{r}
# create a random intercept and random slope model with the interaction
BPPS_ref2 <- lmer(bprs ~ week + treatment + (week | subject) + week*treatment, data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPPS_ref2)

# perform an ANOVA test on the two models
anova(BPPS_ref2, BPPS_ref1)

# draw the plot of BPRSL with the observed bprs values
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(BPPS_ref2)

# Create a new column fitted to RATSL
BPRSL$Fitted <- Fitted

# draw the plot of RATSL with the Fitted values of weight
ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
  geom_line() +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs") +
  theme(legend.position = "top")

```  
  
  